Linear and Differential Cryptanalysis of DES

### 2009-02-02 trunk/talk/draft1.tex
Completed talk ... Final PDF included ... Some more figures ...

\vspace{5mm}
\pause Differential Characteristic
\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]
\end{column}
\end{columns}
\end{frame}

@@

\vspace{5mm}
Constructing the Difference Distribution tables for the S-Boxes and analyzing the probability distribution for differential propagation through SBoxes:

@@

\vspace{3mm}
\pause Differential Characteristic: $20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02$ holds with probability
\[ p_D \: = \: (10/64) \times (16/64) \times (12/64) \times (12/64) \: = \: 45/32768 \approx 0.001373 \]

\vspace{2mm}
\pause Note: $p_D$ is much higher than the random distribution probability $1/2^{64}$ and we hope to exploit this situation to break DES.

\begin{column}{0.6\textwidth}
Differential Characteristic of DES
\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]

\pause Note: The right half of the round 3 characteristic output is $00\:00\:80\:02$.

\vspace{2mm}
\pause Thus, the right halves of the ciphertexts at this round differ by 2 bits, and these are related to S-Box $S_1$ in round 4.

\vspace{2mm}
\pause Hence, we just need to guess the {\it target partial subkey} $K_{4,1}, ..., K_{4,6}$ for $S_1$, i.e. the first 6 bits of $K_4$.

\vspace{2mm}
\pause Thereafter, we can ``undo'' round 4 using this guess and  the maximum probability of satisfying the 3 round differential characteristic will indicate a correct guess.
\end{column}
\end{columns}
\end{frame}

@@

\subsection{Independence of S-Boxes}
\begin{frame}
Note that throughout this discussion we have assumed that the S-Boxes were independent. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds. \\
\vspace{3mm}
This assumption worked well for us in practice, but is not necessarily true. We could have proceeded differently without this assumption - [see John Manferdelli's Boolean Functions slides]
\end{frame}

\subsection{Links between the Attacks}
\begin{frame}
Linear Resistance:\\
\vspace{3mm}
Let $F: GF(2)^p \rightarrow GF(2)^q$.\\
\vspace{3mm}
Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
\vspace{3mm}
Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
\vspace{3mm}
Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
\vspace{5mm}
To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
\end{frame}

\subsection{Take Home}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
`Take Home'
\end{beamercolorbox}
\end{frame}

\begin{frame}
If you are mounting the attacks ... To Do / Not To Do ...

@@

\begin{frame}
The problems of DES originated in two aspects:
\begin{itemize}
\item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
\item{Diffusion between rounds is just not fast enough}
\end{itemize}
AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
\end{frame}

@@

\begin{frame}
\begin{center}
The End.\\
Questions?
\end{center}
\end{frame}

>>>>

\vspace{5mm}
\pause Differential Characteristic
\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]

\end{column}
\end{columns}
\end{frame}

\begin{column}{0.6\textwidth}
\vspace{2mm}
\pause Note: $00\:00\:80\:02$ implies difference of 2 bits which are both related to S-Box $S_1$ in round 4.

\vspace{2mm}
\pause Hence, in round 4
\[ D' = \Delta C_L \oplus 00\:00\:80\:02 \]
\pause and hence for $S_2, ... , S_8$, outputs are
\[ \Delta Y = P^{-1}(\Delta C_L) = S_{box}(E(C_R) \oplus K_4) \]

\vspace{2mm}
So, just guess the {\it target partial subkey} for $S_2, ... S_8$ individually, \pause and compute their predicted $\Delta Y$ with a number of plaintext pairs.

\vspace{2mm}
\pause This will ``undo'' round 4 and  the maximum probability of satisfying the 3 round differential characteristic will indicate a correct guess.
\end{column}
\end{columns}
\end{frame}

@@

\vspace{3mm}
Constructing the Difference Distribution tables for the S-Boxes and analyzing the probability distribution for differential propagation through SBoxes:

@@

\vspace{3mm}
\pause Differential Characteristic for $\Delta C_L$: $00\:00\:00\:00 \rightarrow 00\:00\:80\:02$ holds with probability
\[ p_D \: = \: 10/64 \: \approx \: 0.16 \: \gg 1/2^{32} \]

\vspace{2mm}
\pause
\begin{itemize}[<+->]
\item{How many pairs do we need?}
\item{Approximately $2^4 = 16$ for $100\%$ success}
\end{itemize}

\pause For 16 round DES, $2^{47}$ pairs are needed. [$2^{36}$ ``good pairs'']
\end{frame}

@@

\subsection{A few important points}
\begin{frame}
Independence of S-Boxes

\vspace{5mm}
\pause Note that throughout this discussion we have assumed that the S-Boxes were independent. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds.

\vspace{3mm}
\pause This assumption worked well for us in practice, but is not necessarily true. We could have proceeded differently without this assumption - [see John Manferdelli's Boolean Functions slides]
\end{frame}

\begin{frame}
Linear Resistance

\vspace{5mm}
Let $F: GF(2)^p \rightarrow GF(2)^q$.\\
\vspace{3mm}
\pause Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
\vspace{3mm}
\pause Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
\vspace{3mm}
\pause Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
\vspace{5mm}
\pause To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
\end{frame}

\subsection{Take Home}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
`Moral of the story'
\end{beamercolorbox}
\end{frame}

\begin{frame}
If you are mounting the attacks, remember the following:

\vspace{5mm}
\pause
\begin{itemize}[<+->]
\item{Analyze the main non-linear component, S-Boxes for example}
\item{For linear attack, try to find the linear/affine approximation}
\item{For differential attack, find the best characteristic}
\item{Combine through levels to get the total approximation/characteristic}
\item{Analyze piling up probabilities to estimate number of required pairs}
\item{Verify if this is better than any other existing attack}
\end{itemize}

@@

\begin{frame}
If you are designing a block cipher, note:

\vspace{5mm}
\pause The problems of DES originated in two aspects
\begin{itemize}[<+->]
\item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
\item{Diffusion between rounds is just not fast enough}
\end{itemize}

\vspace{2mm}
\pause Prescription
\begin{itemize}[<+->]
\item{Create S-Boxes as bent functions [inefficient as $p \geq 2q$]}
\item{Make the diffusion step faster to create an avalanche}
\end{itemize}

\vspace{3mm}
\pause AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
\end{frame}

\subsection{Fun facts}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Just for Fun!
\end{beamercolorbox}
\end{frame}

\begin{frame}
What if you play around with the S-Box ordering?

\vspace{5mm}
\pause
\begin{itemize}[<+->]
\item{Matsui prescribed a rearrangement of S-Boxes specifically against linear attack.}
\item{Matsui prescribed another rearrangement of the S-Boxes, which is good enough to prevent both linear and differential attacks.}
\item{Biham and Shamir illustrated a rearrangement where differential attack becomes way more efficient.}
\end{itemize}

\begin{frame}
\begin{figure}
\includegraphics[width=0.6\textwidth]{ibm.jpg}
\end{figure}

\pause
S-Box Design criteria
\begin{itemize}[<+->]
\item{1 bit input difference produces 2 bits output difference}
\item{Minimize the difference between the numbers of 1's and 0's when any input bit is held constant}
\item{$S(X) \neq S(X \oplus 11ab00)$} 
\end{itemize}

\vspace{2mm}
\pause Other good stuff
\begin{itemize}[<+->]
\item{2 input difference bits mapped to 3 by $E(R)$}
\item{Each output difference bit permuted to 1 each to S-boxes of next round}
\end{itemize}

\begin{frame}
\begin{figure}
\includegraphics[totalheight=0.7\textheight]{coppersmith.jpg}
\end{figure}
\pause
\centerline{Average of $10$ and $10^{45}$ is $5.10^{44} + 5 > 10^{44} \approx 2^{47}$}
\pause
\begin{figure}
\includegraphics[width=0.7\textwidth]{eli.jpg}
\end{figure}
\hfill - Eli Biham
\end{frame}

@@

\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
The End.
\end{beamercolorbox}
\end{frame}

\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Questions?
\end{beamercolorbox}
\end{frame}
